---
title: "Barbell Prediction"
author: "Fábio Franco Costa"
date: "17/05/2015"
output: html_document
---
## Executive Summary
In this assignment we created a model to predict 5 possible *classes*. The final model was tested in 20 cases supplied by the course staff. Our model have a very high accuracy and we managed to correctly predict every case.

## Model creation
For this assignment we created a model from a subset of the original dataset.  

```{r, echo=FALSE}
# Read Data
pmlData <- read.csv("pml-training.csv")
pmlFinalTest <- read.csv("pml-testing.csv") 

# Copy data to avoid reload the CSV. Not really necessary but usefull in 
# development. Pardon my lazyness to set cache=FALSE ;-) 
data <- pmlData
classe <- data$classe
```

We dropped columns with variables that should not be included in the model like **'user_name'**. We worried that some of this variables could bias the model. For instance, the user could bias the analysis should one of then exercise in a more correct or incorrect way. We also removed some variables that are clearly full of NAs from simple visual inspection. 
```{r}
# Drop unnecessary columns
drops <- c('X', 'user_name', 'raw_timestamp_part_1', 'raw_timestamp_part_2',
         'cvtd_timestamp','new_window', 'num_window', 'kurtosis_yaw_belt',
         'skewness_yaw_belt', 'kurtosis_yaw_dumbbell', 'skewness_yaw_dumbbell',
         'amplitude_yaw_dumbbell', 'amplitude_yaw_belt')
data <- data[,!(names(data) %in% drops)]
```

Many numeric variables were imported as factor. We managed to correct that by creating two auxiliary functions as shown below:
```{r, warning=FALSE}
# Auxiliary functions
asNumeric <- function(x) as.numeric(as.character(x))
factorsNumeric <- function(d) modifyList(d, lapply(d[, sapply(d, is.factor)], 
                                                   asNumeric))
# Convert variables from factor
data <- factorsNumeric(data)
```

Due to factor/numeric conversion some values were coerced to NAs. We decided to remove columns that have some NAs. In a more complex setting we could try to impute this values, but our exercise did not need this extra processing.
```{r}
# Remove columns that have NAs
data <- data[,colSums(is.na(data)) == 0]
```
```{r, echo=FALSE}
# Add classe variable back
data <- data.frame(classe, data)
```

We will use the **Caret** package to create our model. The first step is data partitioning as training and testing datasets.
```{r, echo=FALSE, message=FALSE}
# Load packages and set seed
library("caret")
library('scales')
set.seed(666)
```
```{r}
# Create trainning datasets
inTrain <- createDataPartition(y = data$classe, p = 0.8, list = F)
training <- data[inTrain,]
testing <- data[-inTrain,]
```

We train the model, with a 10-fold cross validation strategy. You can see bellow that we enclosed our train function in **system.time()**. We used execution time as benchmark to fine tune some options during development. 
```{r, cache=TRUE}
# Setup model parameters
train_control <- trainControl(## 10-fold CV
                           method = "cv",
                           number = 10,
                           allowParallel=TRUE)

# Fit model: random forest
system.time({
modelFit <- train(classe ~ .,
                  data = training,
                  trControl=train_control, 
                  verbose=FALSE,
                  method=c("rf"))
})
```

Finally, we can use the **confusionMatrix()** to check our predictions and get the **Accurary** from out model. We will use that to estimate the **Error Rate**.
```{r, echo=FALSE}
# Test the model
predictions <- predict(modelFit, newdata = testing)
CM <- confusionMatrix(predictions, testing$classe)
error <- 1 - CM$overall[1]
CM
```
## Cross Validation and Error
For cross validation, we adopted a K-Fold strategy with 10 folds. We used **trainControl()** to set our options.  
Our out of sample error is estimated to be `r percent(error)`

## Test Cases
Our prediction model was able to predict correctly all the assignments test cases.
```{r, echo=FALSE}
# Predict real cases
# Output ommited for final report
cases <- predict(modelFit, newdata = pmlFinalTest)

# Generate predictions
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
#setwd("~/Work/DSCoursera/datasciencecoursera/barbell/results")
#pml_write_files(cases)
#https://www.youtube.com/watch?v=hifQl_9Ugqo
```
